Utility of Diffusion of Innovations for Diagnosing the Spread of Conservation Behaviors

### 2021-12-19 DOI_Review/Manuscript.rmd
Kelly edits & added Jeff

---
title: Utility of Diffusion of Innovations for Diagnosing the Spread of Conservation Behaviors
author:
- affiliation: Boise State University - Human Environment Systems
  name: Matt Clark
- affiliation: Boise State University- Human Environment Systems
  name: Vicken Hillis

@@

Decades of research have identified key contextual factors that support the durability of resource safeguarding behaviors in a population, particularly with regard to common-pool resources [@agrawal_common_2001; @baland_halting_1996; @ostrom_governing_1990]. These factors are largely narrative descriptions of the qualities of, and interactions between, resource types, political structures, and resource users which are most likely to lead to stable and sustainable resource management institutions [@ostrom_general_2009]. Although at least 40 unique factors have been identified as contributing to the success of community-based conservation institutions, well defined social and ecological boundaries, and well connected social systems stand out as major important themes [@ostrom_governing_1990; @ostrom_general_2009; @agrawal_common_2001]. Research on common property additionally recommends that institutions be matched in scale to the underlying ecological processes of the resources they aim to manage [@mckean_common_2000].

@@

Diffusion of novel beliefs and behaviors is a specific instance of cultural evolution, where the trait being transmitted is new to the population and outcompetes some previous cultural variant, such as in the classic example of hybrid corn seed adoption in Iowa farmers [@mesoudi_cultural_2016; @ryan_acceptance_1950]. We observe these patterns using diffusion curves which show the proportion of a population that has adopted a belief or behavior on the Y-axis, as a function of time on the X-axis [@hoppitt_effect_2010; @rogers_diffusion_2010; @ryan_acceptance_1950]. The exact shape of these diffusion curves is highly dependent on the learning mechanism employed by individuals in the population, as well as contextual factors like the population density or ease of trialing the cultural variant [@smaldino_adoption_2017; @henrich_cultural_2001; @rogers_diffusion_2010].

@@

We can also use these parameter values, estimated from the first 30% of the time series, to predict the long-term patterns of adoption for a community-based conservation initiative. In figures 5 & 6, the translucent black lines show 1000 draws from the posterior predicting the proportion of the population who will be enrolled in the intervention at every time step. Gray and blue points show the initial 30% of the data which was used to estimate the three parameters of interest and the remaining 70% of data respectively. These figures show that the SIRS model accurately captures and predicts the pattern of behavioral adoption for both unbiased learning and environmentally mediated success-biased learning. R and STAN code for these models can be found in the supplemental material (S3).

@@

Mechanistic, or process models allow for us to formally describe these complex interactions while testing the logical consequences of our assumptions and the relation of our hypothesized processes to observed data [@epstein_why_2008]. The practice of applying statistical frameworks for parameter estimation to analytical process models, and in particular to ordinary differential equations, is however still relatively new and has been embraced mostly in ecology [@hefley_when_2017; @laubmeier_ecological_2020; @morales_mechanistic_2021]. This emerging methodology allows for the estimation of interesting and directly interpretable process parameters, not possible with linear regression. We believe this is an exciting new avenue for conservation science as the direct interpretability and applicability of process parameters lend themselves well to tight integration with applied research as is common in the field. Future research may be interested in extending the SIRS diffusion model to include a term which represents change in the environment over time and the effect of resource levels on adoption rates to more accurately reflect the social-ecological coupling. An example of such an extension is presented below, where we include another equation for the change in a resource pool over time $\frac{dE}{dt}$. We also must include three additional parameters: $\phi$, $\rho$, and $\theta$, which represent the effect of resource level on enrollment (infection), the effect of harvest on the resource level, and the regeneration rate of the resource respectively.

@@

Coupled human and natural systems commonly exhibit dynamics with thresholds, resilience, and heterogeneity not present in social or natural systems in isolation and therefore are difficult to model analytically [@liu_complexity_2007]. In very complex cases such as these, we argue that agent-based simulations are the preferred approach as they allow for these features of complex systems which are difficult or even impossible to account for with process models such the SIRS  or SIRS extension model presented in equation 4 [@reyers_social-ecological_2018; @railsback_agent-based_2012; @wilensky_introduction_2015]. Agent-based simulations allow modelers to impose a realism in the governance or institutional dynamics that are largely intractable in closed-form mathematical equations [@bravo_agents_2011; joffre_combining_2015]. As the number of purely subsistence communities shrinks each year worldwide, and multiple levels of government are increasingly involved in natural resource management, incorporating these features into our models is paramount for reliable inference [@mckean_common_2000; @armitage_adaptive_2009].

>>>>

---
title: Utility of Diffusion of Innovations for Diagnosing the Spread of Conservation Behaviors
author:
- affiliation: Boise State University - Human Environment Systems
  name: Matt Clark
- affiliation: Boise State University - Human Environment Systems
  name: Vicken Hillis
- affiliation: Max Planck Institute for Evolutionary Anthropology 
  name: Jeffrey Andrews

@@

Decades of research have identified key contextual factors that support the durability of resource safeguarding behaviors in a population, particularly with regard to common-pool resources [@agrawal_common_2001; @baland_halting_1996; @ostrom_governing_1990]. These factors are largely narrative descriptions of the qualities of, and interactions between, resource types, governance structures, and resource users which are most likely to lead to stable and sustainable resource management institutions [@ostrom_general_2009]. Although at least 40 unique factors have been identified as contributing to the success of community-based conservation institutions, well defined social and ecological boundaries, and well connected social systems stand out as major important themes [@ostrom_governing_1990; @ostrom_general_2009; @agrawal_common_2001; @cox_review_2010]. Research on common property additionally recommends that institutions be matched in scale to the underlying ecological processes of the resources they aim to manage [@mckean_common_2000].

@@

Diffusion of novel beliefs and behaviors is a specific instance of cultural evolution, where the trait being transmitted is new to the population and outcompetes some previous cultural variant, such as in the classic example of hybrid corn seed adoption in Iowa farmers [@mesoudi_cultural_2016; @ryan_acceptance_1950]. We observe these patterns using diffusion curves which show the proportion of a population that has adopted a belief or behavior on the Y-axis, as a function of time on the X-axis [@hoppitt_effect_2010; @rogers_diffusion_2010; @ryan_acceptance_1950]. The exact shape of these diffusion curves can be highly dependent on the learning mechanism employed by individuals in the population, as well as contextual factors like the population density or ease of trialing the cultural variant [@smaldino_adoption_2017; @henrich_cultural_2001; @rogers_diffusion_2010].

@@

We can also use these parameter values, estimated from the first 30% of the time series, to predict the long-term patterns of adoption for a community-based conservation initiative. In figures 5 & 6, the translucent black lines show 1000 draws from the posterior predicting the proportion of the population who will be enrolled in the intervention at every time step. Gray and blue points show the initial 30% of the data which was used to estimate the three parameters of interest and the remaining 70% of data, respectively. These figures show that the SIRS model accurately captures and predicts the pattern of behavioral adoption for both unbiased learning and environmentally mediated success-biased learning. R and STAN code for these models can be found in the supplemental material (S3).

@@

Mechanistic, or process models allow for us to formally describe these complex interactions while testing the logical consequences of our assumptions and the relation of our hypothesized processes to observed data [@epstein_why_2008]. The practice of applying statistical frameworks for parameter estimation to analytical process models, and in particular to ordinary differential equations, is however still relatively new and has been embraced mostly in ecology [@hefley_when_2017; @laubmeier_ecological_2020; @morales_mechanistic_2021]. This emerging methodology allows for the estimation of interesting and directly interpretable process parameters, not possible with linear regression. We believe this is an exciting new avenue for conservation science as the direct interpretability and applicability of process parameters lend themselves well to tight integration with applied research as is common in the field. Future research may be interested in extending the SIRS diffusion model to include a term which represents change in the environment over time and the effect of resource levels on adoption rates to more accurately reflect the social-ecological coupling. An example of such an extension is presented below, where we include another equation for the change in a resource pool over time $\frac{dE}{dt}$. We also must include three additional parameters: $\phi$, $\rho$, and $\theta$, which represent the effect of resource level on enrollment (infection), the effect of harvest on the resource level, and the regeneration rate of the resource, respectively.

@@

Coupled human and natural systems commonly exhibit dynamics with thresholds, resilience, and heterogeneity not present in social or natural systems in isolation and therefore are difficult to model analytically [@liu_complexity_2007]. In very complex cases such as these, we argue that agent-based simulations are the preferred approach as they allow for these features of complex systems which are difficult or even impossible to account for with process models such the SIRS  or SIRS extension model presented in equation 4 [@reyers_social-ecological_2018; @railsback_agent-based_2012; @wilensky_introduction_2015]. Agent-based simulations allow modelers to impose a realism in the governance or institutional dynamics that are largely intractable in closed-form mathematical equations [@bravo_agents_2011; joffre_combining_2015]. As the number of purely subsistence communities shrinks each year worldwide, and multiple levels of institutions are increasingly involved in natural resource management, incorporating these features into our models is paramount for reliable inference [@mckean_common_2000; @armitage_adaptive_2009].
